---
title: "test"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{user-guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*This article does not contain any real patient data. All patient data has been simulated but formatted to match the structure of CPRD Aurum data.*



# Introduction

This workflow is advantageous as it avoids hard file paths, beneficial if wanting to move your code onto another computer system. Furthermore, once codelists and SQLite database have been created and stored in the appropriate folder, they can simply be referred to by name, resulting in an easier workflow. The function `create_directory_system()` will create the directory system required to use rAURUM in this way. To avoid repetition of the previous section, this is showcased just once using the `extract_ho` function. We start by creating a temporary directory, setting this to be the working directory, and generating the required directory structure:

```{r}
library(rAURUM)
getwd()

## Create directory system compatible with rAURUM's automatic saving of output
create_directory_system()
```

Next we create a SQLite database called "mydb.sqlite" in the "data/sql" directory, and some data to this. Note this document is generated with Rmarkdown, so the working directory must be set every time in the next couple of code chunks.

Next we create and save a code list called "mylist.csv" into the "codelists/analysis/" directory.

```{r}
getwd()
```
